Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North

### 2024-02-01 reports/GnC_report/paper/paper.qmd
update report itself

```{r, setup, include=FALSE}
root_dir <- here::here()
paper_dir <- here::here("reports", "GnC_report", "paper")
knitr::opts_knit$set(root.dir = root_dir)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
backup_options <- options()
options(scipen = 1, digits = 2)
do_eval <- FALSE
```

@@

@@

```{r echo=FALSE}
#| label: spatial-basics

RoF_CameraDeploymentLog_WLU <- 
  readxl::read_excel(paste0("data-raw/tabular/RoF_CameraDeploymentLog.xlsx")) |> 
  janitor::clean_names()

cams_sf <- RoF_CameraDeploymentLog_WLU |> 
  dplyr::rename(longitude = x, latitude = y) |> 
  dplyr::filter(!is.na(latitude)) |> 
  dplyr::filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude","latitude"), crs = "EPSG:4326", remove = F) |> 
  dplyr::mutate(is_retrieved = ifelse(!is.na(date_retrieved), "Yes", "No"))

coords <- cams_sf |> 
  sf::st_coordinates() |> 
  as.data.frame()

cams_sf_coords <- dplyr::bind_cols(cams_sf, coords)

zones <- 
  sf::st_read("data-raw/spatial/ecozones/EcoZone/EcoZone.shp",
              quiet = TRUE, ) |> 
  sf::st_transform(crs = sf::st_crs(cams_sf))

zones_filter <- zones |>
  sf::st_filter(cams_sf)

pols <- sf::st_coordinates(zones_filter) |> 
  as.data.frame()

regions_labels <- zones_filter |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.5),
                              y = c(52, 52.6)))
```

```{r eval=do_eval, echo=FALSE}
#| label: spatial-compute

map <- ggmap::get_stadiamap(bbox = c(left = -90, bottom = 50, right = -83, top = 55.5), 
                            zoom = 5, maptype = "stamen_terrain")

base <- ggmap::ggmap(map) +
  ggplot2::labs(x = "Longitude", y = "Latitude") +
  ggplot2::geom_path(data = pols, ggplot2::aes(x = X, y = Y), col = 1, linewidth = 0.2)
saveRDS(base, here::here(paper_dir, "base_map.rds"))

p <- base +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords,
                      shape = 24, col = "black", fill = "red",
                      ggplot2::aes(x = X, y = Y), 
                      inherit.aes = FALSE)
ggplot2::ggsave(plot = p, filename = "figures/camera_array.png", 
                width = 6, height = 8, dpi = 300)
saveRDS(p, here::here(paper_dir, "camera_array.rds"))
```

```{r echo=FALSE}
#| label: fig-deployment
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."
p <- readRDS(here::here(paper_dir, "camera_array.rds"))
p
```

```{r eval=do_eval, echo=FALSE}
#| label: fig-deployment-poster

@@

```{r eval=do_eval, echo=FALSE}
#| label: process-exif

exif <- readr::read_csv("data/tabular/all_exif.csv") |>
  janitor::clean_names() |>
  dplyr::mutate(image_id = stringr::str_replace_all(source_file, "/", "_")) |>
  dplyr::mutate(image_id = stringr::str_replace_all(image_id, ".JPG", "")) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage/")) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/workspace/storage/TrailCamStorage_2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/workspace/storage/TrailCamStorage/")) |> 
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage_")) |> 
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_workspace_storage_TrailCamStorage_2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_workspace_storage_TrailCamStorage_")) |> 
  dplyr::mutate(image_id = strsplit(image_id, "_") |> 
                  purrr::map(~paste0(.x[2:length(.x)], collapse = "_")) |> 
                  unlist()) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage/"))) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage_2/")))

readr::write_csv(exif, here::here("data", "tabular", "exif_processed.csv"))
```

```{r echo=FALSE}
#| label: load-data

anns_wide <- readRDS("data/objects/annotations_wide_treated.rds")
anns_wide_noempty  <- readRDS("data/objects/annotations_wide_noempty.rds")
dets <- readRDS("data/objects/detections.rds")
exif <- readr::read_csv(here::here("data", "tabular", "exif_processed.csv"))

@@

```{r echo=FALSE}
#| label: mans

@@

We summarized our existing species data to date for analysis in two main ways:  (1) species presence/absence data for each camera and, (2) relative species abundance, measured as the number of days, within each month that each species was detected at a camera site.

```{r echo=FALSE}
#| label: quality-control

# Only annotations for animals
# TODO investigate id of cut length 1
anns_wide_animals <- anns_wide_noempty |>
  dplyr::filter(species != 'Human') |> 
  dplyr::filter(species != 'Vehicle') |> 
  dplyr::mutate(id_cut_length = purrr::map_dbl(id , \(x) {
    x |> stringr::str_split("_") |> purrr::map_dbl(~length(.x))
  })) |> 
  dplyr::filter(id_cut_length > 1) |> 
  dplyr::mutate(camera_id = purrr::map_chr(id , \(x) {
    x |>
      stringr::str_split("_") |>
      purrr::map(~.x[[2]]) |>
      unlist()
  }))

# Unknowns and bird breakdown
# anns_wide_animals |> dplyr::filter(species == "Unknown") |> 
#   dplyr::pull(id) |> 
#   stringr::str_sub(1,4) |> table() |> sort() 
# anns_wide_animals |> dplyr::filter(stringr::str_detect(species,"Bird")) |> 
#   dplyr::pull(id) |> 
#   stringr::str_sub(1,4) |> table() |> sort()

# Joins

# TODO streamline exif so you dont have to join
anns_wide_animals_joined <- anns_wide_animals |>
  dplyr::left_join(exif, by = c("source_file", "image_id"))

stopifnot(dim(anns_wide_animals)[1] ==  dim(anns_wide_animals_joined)[1])
stopifnot((anns_wide_animals_joined$source_file |> is.na() |> sum()) == 0)

anns_wide_animals_joined_mut <- anns_wide_animals_joined |> 
  dplyr::filter(species != "Unknown") |>
  dplyr::mutate(plot_id = stringr::str_sub(id, 1, 4)) |> 
  dplyr::mutate(date_time = lubridate::ymd_hms(exif_date_time_original,
                                               tz = "Canada/Eastern"),
                year = lubridate::year(date_time),
                month = lubridate::month(date_time, label = TRUE),
                day = lubridate::day(date_time),
                ymd = paste(year, month, day, sep = "_"))

# anns_wide_animals_joined_dets <- anns_wide_animals |> 
#   dplyr::right_join(dets_images, by = "image_id")

saveRDS(anns_wide_animals_joined_mut, "data/objects/annotations_animals.rds")
```

<!-- Other camera plots -->

```{r eval=FALSE, echo=FALSE}
# library(forcats)
library(ggplot2)
# anns_wide_animals_joined_mut |> 
#   ggplot(aes(x=date_time, y=species)) + geom_point()
# anns_wide_animals_joined_mut |> 
#   dplyr::filter(species == "Fisher") |> View()

camera_days <- exif |> dplyr::group_by(maker_notes_user_label) |> 
  dplyr::summarise(camera_first_time=min(exif_date_time_original)) |> 
  dplyr::mutate(date_time = lubridate::ymd_hms(camera_first_time,
                                               tz = "Canada/Eastern"),
                year = lubridate::year(date_time),
                month = lubridate::month(date_time, label = TRUE),
                day = lubridate::day(date_time),
                ymd = paste(year, month, day, sep = "_")) |> 
  dplyr::select(camera_id = maker_notes_user_label, 
                camera_first_day = date_time)

camera_regions <- cams_sf |> sf::st_join(zones) |> 
  sf::st_drop_geometry() |> dplyr::select(camera_id, ECOZONE_NA)

bp <- anns_wide_animals_joined_mut |>

  dplyr::filter(species != "Fisher") |> 
  dplyr::left_join(camera_days, by = "camera_id") |>

@@

  dplyr::mutate(diff_time = difftime(first_day, camera_first_day, units="days"),
                ECOZONE_NA = as.factor(ECOZONE_NA)) |>

  ggplot(aes(x = diff_time, 
             y = forcats::fct_reorder(species, total_events),
             color = ECOZONE_NA)) +
  geom_boxplot() +

  scale_x_continuous()+
  labs(x = "Detection time (in days)",
       y = "Species",
       color = "Ecozone") + 
  theme_bw()

ggsave(plot = bp, filename = "figures/bp.png", width = 10, height = 15)

pts <- anns_wide_animals_joined_mut |>

  dplyr::filter(species != "Fisher") |> 
  dplyr::left_join(camera_days, by = "camera_id") |>

@@

  dplyr::mutate(diff_time = difftime(first_day, camera_first_day, units="days"),
                ECOZONE_NA = as.factor(ECOZONE_NA)) |>

  ggplot(aes(x = diff_time, 
             y = forcats::fct_reorder(species, total_events),
             # size = camera_events, 
             color = ECOZONE_NA)) +

  geom_point(aes(x = diff_time, 
                  y = forcats::fct_reorder(species, total_events),
                  size = camera_events, 
                  color = ECOZONE_NA,
                  group = ECOZONE_NA)) +

  scale_x_continuous()+
  labs(x = "Detection time (in days)",
       y = "Species",
       color = "Ecozone",
       size = "Day counts") + 
  theme_bw()
ggsave(plot = pts, filename = "figures/pts.png", width = 10, height = 15)
pts
```

```{r eval=FALSE, echo=FALSE}
# for_vic <- anns_wide_animals_joined_mut |> 
#   dplyr::select(id,species,date_time, camera_id) |> 
#   dplyr::left_join(cams_sf |> dplyr::select(camera_id, plot_id) |> sf::st_drop_geometry(),
#                    by = "camera_id")
# tot_camera <- for_vic$camera_id |> unique() |> length()
# hit_rates <- for_vic |> dplyr::select(species, camera_id) |> 
#   dplyr::group_by(species) |> dplyr::summarise(n_cameras=length(unique(camera_id))) |> 
#   dplyr::ungroup() |> dplyr::mutate(rate = n_cameras/tot_camera*100) |> 
#   dplyr::arrange(dplyr::desc(rate))
# # det_history_two_weeks <- 
# 
# readr::write_csv(for_vic, "../data/derived_data/observations_by_camera_2023Aug15.csv")
# readr::write_csv(hit_rates, "../data/derived_data/hit_rates_by_species_2023Aug15.csv")
```

```{r echo=FALSE}
#| label: species

anns_wide_animals_table <- anns_wide_animals |>
  dplyr::select(camera_id, species) |> 
  dplyr::filter(species != "Unknown") |>
  table() 
anns_wide_animals_table <- anns_wide_animals_table > 0
anns_wide_animals_table <- anns_wide_animals_table + 0
anns_wide_animals_table <- anns_wide_animals_table |> 
  as.data.frame() |> 
  tibble::rownames_to_column("camera_id") |> 
  dplyr::arrange(camera_id)
anns_wide_animals_table_tb <- anns_wide_animals_table |> 
  dplyr::select(-camera_id)
rownames(anns_wide_animals_table_tb) <- anns_wide_animals_table$camera_id
```

@@

```{r}
#| label: extract
#| echo: false

lu_16 <- raster::raster("data-raw/spatial/land_use/Class/FarNorth_LandCover_Class_UTM16.tif")
lu_17 <- raster::raster("data-raw/spatial/land_use/Class/FarNorth_LandCover_Class_UTM17.tif")

lu_dat <- readr::read_csv("data-raw/spatial/land_use/attr_table_northen_ont_lc.txt") |> 
  dplyr::mutate(cats = as.factor(code))

cams_sf_buffers_16 <- cams_sf |> 
  sf::st_transform(sf::st_crs(lu_16)) |> 
  sf::st_buffer(dist = 1000)

cams_sf_buffers_17 <- cams_sf |> 
  sf::st_transform(sf::st_crs(lu_17)) |> 
  sf::st_buffer(dist = 1000)

extr_16 <- exactextractr::exact_extract(lu_16, cams_sf_buffers_16,
                                        progress = FALSE)
extr_17 <- exactextractr::exact_extract(lu_17, cams_sf_buffers_17,
                                        progress = FALSE)

ind_16 <- which(lapply(extr_16, function(x) all(is.na(x[,1]))) |> unlist())
ind_17 <- which(lapply(extr_17, function(x) all(is.na(x[,1]))) |> unlist())

stopifnot(length(unique(c(ind_16, ind_17))) == length(extr_16))

extr_16[ind_16] <- extr_17[ind_16]

extr <- extr_16

extr_list <- extr |> 
  lapply(FUN = function(x){
    the_tab <- table(x$value)
    props <- the_tab/sum(the_tab)
    cats <- names(the_tab)
    df <- tibble::tibble(cats, props)
  })
names(extr_list) <- cams_sf_buffers_16$camera_id

extr_table <- dplyr::bind_rows(extr_list, .id = "camera_id") |> 
  dplyr::mutate(props = as.double(props)) |> 
  dplyr::left_join(lu_dat, by="cats") |> 
  dplyr::filter(!is.na(category_code)) |> 
  dplyr::select(camera_id,category_code, props) |> 
  tidyr::pivot_wider(names_from = "category_code", values_from = "props") |> 
  as.data.frame()
extr_table[is.na(extr_table)] <- 0

# Combine the categories
extr_table_simple <- extr_table |> 
  dplyr::mutate(WAT = WAT + XWAT,
                MixTRE = MixTRE + DecTRE,
                Dist  = TrOrSHr + NSWood, 
                SWA = ThSWA + ConSWA, 
                Anth = URB + MIN + BED) |> 
  dplyr::select(-XWAT,
                -DecTRE,
                -TrOrSHr, -NSWood, 
                -ThSWA, - ConSWA, 
                -URB, -MIN, -BED)

extr_table_sub <- extr_table_simple |> 
  dplyr::filter(camera_id %in% anns_wide_animals_table$camera_id) |> 
  dplyr::arrange(camera_id)
extr_table_sub_tb <- extr_table_sub |> 
  dplyr::select(-camera_id)
rownames(extr_table_sub_tb) <- extr_table_sub$camera_id
extr_table_sub_tb <- extr_table_sub_tb[, colSums(extr_table_sub_tb) != 0]

# lu_cropped <- raster::crop(lu_16, cams_sf_buffers_16)
# raster::plot(lu_cropped)
```

@@

```{r}
#| label: events-Plots
#| echo: false

events_plot_days <- anns_wide_animals_joined_mut |> 
  # dplyr::mutate(ind = paste(species, sex, age, sep = "_")) |> 
  dplyr::select(plot_id, species, ymd) |> 
  dplyr::distinct()

events_plot <- events_plot_days |> 
  dplyr::group_by(plot_id, species) |>
  dplyr::summarise(day_count=dplyr::n()) |> 
  dplyr::ungroup()

events_plot_tb <- events_plot |> 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") |>
  dplyr::arrange(plot_id) |>
  tibble::column_to_rownames("plot_id")

events_plot_tb[is.na(events_plot_tb)] <- 0
events_plot_tb_filt <- events_plot_tb[,colSums(events_plot_tb)>50]
tmp_col <- ncol(events_plot_tb_filt)

events_plot_tb_filt_with_col <- events_plot_tb_filt |> 
  tibble::rownames_to_column("plot_id")
```

```{r}
#| label: events-Cams
#| echo: false

events_cam_days <- anns_wide_animals_joined_mut |> 
  dplyr::select(camera_id, species, month, ymd) |> 
  dplyr::distinct()

events_cam <- events_cam_days |> 
  dplyr::group_by(camera_id, species) |>
  dplyr::summarise(day_count=dplyr::n()) |> 
  dplyr::ungroup()

events_tb_cam <- events_cam |> 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") |>
  dplyr::arrange(camera_id) |>
  tibble::column_to_rownames("camera_id")

events_tb_cam[is.na(events_tb_cam)] <- 0
# events_tb_cam_filt <- events_tb_cam[,colSums(events_tb_cam)>50]
# tmp_col <- ncol(events_tb_cam)
```

```{r}
#| label: tab-species-table
#| echo: false

sp_tab <- dplyr::rename(.data = events_tb_cam |> colSums() |> 
                          sort(decreasing = TRUE) |> as.data.frame(), 
                        days_detected=`sort(colSums(events_tb_cam), decreasing = TRUE)`) |> 
  tibble::rownames_to_column("species")

# sci_names <- taxize::comm2sci(sp_tab$species)
# saveRDS(sci_names, "analysis/data/derived_data/sci_names.rds")
sci_names <- readRDS("data/objects/sci_names.rds") |> 
  unlist() |> tibble::as_tibble(rownames = NA) |> 
  tibble::rownames_to_column() |> 
  dplyr::rename(Scientific_name = value,
                species = rowname) |> 
  tibble::add_row(Scientific_name = "Ursus americanus", species="Black bear") |> 
  tibble::add_row(Scientific_name = "Perisoreus canadensis", species="Canada Jay")

sp_tab |>
  dplyr::left_join(sci_names, by="species") |> 
  dplyr::relocate(Scientific_name, .after=species) |> 
  dplyr::filter(days_detected >= 10) |>
  gt::gt() |>
  gt::tab_header(
    title = "Table 2: Relative abundances for species with more than 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) |>
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) |> 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
    Scientific_name = "Scientific name"
  ) |> 
  gt::tab_style(style = list(
      gt::cell_text(style="italic")),
    locations = gt::cells_body(columns = Scientific_name)) |> 
  gt::opt_stylize(style = 6, color = "gray")

@@

```{r}
#| label: fig-map-pies
#| echo: false
#| fig-cap: "Pie chart map of relative species abundances across our study area Detections have grouped by plot for better visibility, with a plot containing up to 4 cameras. Only species detected more than 50 days in the season are shown for clarity."
#| fig.width: 8
#| fig.height: 6

cams_sf_with_props <- cams_sf_coords |> 
  sf::st_drop_geometry() |> 
  dplyr::select(plot_id,X,Y) |> 
  dplyr::group_by(plot_id) |> 
  dplyr::mutate(X=mean(X), Y=mean(Y)) |>
  dplyr::distinct() |> 
  dplyr::ungroup() |> 
  dplyr::left_join(events_plot_tb_filt_with_col, by = "plot_id")

# ggplot2::ggplot() +
#   scatterpie::geom_scatterpie(data = cams_sf_with_props,
#                               ggplot2::aes(x=X, y=Y, group=plot_id),
#                               cols=names(cams_sf_with_props)[4:(4+tmp_col-1)]) + 
#   ggplot2::coord_equal() #+ 
#ggplot2::theme(legend.position = "none")

# base2 <- ggplot2::ggplot() +
#   ggplot2::theme_bw()
# for (pol in 1:2){
#   base2 <- base2 + 
#     ggplot2::geom_polygon(data = dplyr::filter(pols, L2 == pol),
#                           ggplot2::aes(x=X, y=Y),
#                           col = 1, fill = "grey90")
# }
# 
# base2 +
#   ggplot2::coord_equal(xlim = c(-88.5, -84),
#                        ylim = c(50, 54)) +
#   scatterpie::geom_scatterpie(data = cams_sf_with_props,
#                               ggplot2::aes(x=X, y=Y, group=plot_id),
#                               cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
#                               sorted_by_radius = TRUE,
#                               legend_name = "Species",
#                               pie_scale = 1.5) +
#   ggplot2::labs(x="Lat", y="Long")

regions_labels_bis <- zones_filter |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-85, -88),
                              y = c(52, 52.6)))

base <- readRDS(here::here(paper_dir, "base_map.rds"))
base +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long") +
    ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels_bis)
```

@@

```{r}
#| label: rda-pa
#| echo: false

source("scripts/R/postprocess/3_ordination.R")

# Ordination
stopifnot(all(rownames(anns_wide_animals_table_tb)==
                rownames(extr_table_sub_tb)))

anns_wide_animals_table_tb_ch <- vegan::decostand(anns_wide_animals_table_tb, 
                                                  "hellinger")

pa_rda <- vegan::rda(anns_wide_animals_table_tb ~ .,
                     extr_table_sub_tb)
```

@@

@@

```{r}
#| label: fig-rda-pa
#| echo: false
#| fig-cap: "Triplot of wildlife presence-absence data from Northern Ontario camera trapping images collected in March through September of 2022 reveals that different species associate with different land cover classes. For example, caribou cranes are more likely to be detected near open and treed wetlands (OBOG, OFEN, TrBOG, and TrFEN) while black bear and moose near forested areas (MixTRE, SpTRE, ConTRE)."
library(magrittr)
custom_rda_plot(pa_rda)
```

@@

```{r}
#| label: rda-events
#| echo: false
#| fig-cap: "Triplot of RDA on relative abundance data"

@@

```{r}
#| label: fig-rda-events
#| echo: false
#| fig-cap: "Triplot of wildlife relative abundance data from Northern Ontario camera trapping images collected in March through September of 2022, showing similar patterns than the triplot using presence-absence data."

@@

# vegan::RsquareAdj(events_rda)
custom_rda_plot(events_rda, sp_scale = 1.5, scale_x_sp=0.4, scale_y_sp=0.3,
                thres_plot = 0.3)
```

@@

```{r}
#| label: cluster-base
#| echo: false

@@

```{r}
#| label: fig-rda-events-clust
#| echo: false
#| fig-cap: "Triplot of wildlife relative abundance data from Northern Ontario camera trapping images collected in March through September of 2022, showing similar patterns than the triplot using presence-absence data. Two clusters determined using Ward clustering are shown as ellipses."

clust_grp <- cutree(clust, k=4)
custom_rda_plot(events_rda, sp_scale = 1.5, scale_x_sp=0.4, scale_y_sp=0.3,
                thres_plot = 0.2716, clust = clust_grp)
```

@@

```{r}
#| label: fig-map-clust
#| echo: false
#| fig-cap: "Map of clustered camera sites. The two clusters were determined using the Ward clustering method."

@@

base_cut <- base +
  ggplot2::coord_map(xlim = c(-89, -84),
                     ylim = c(50, 54.5))

base_cut +
  ggplot2::geom_point(data = cams_sf_clust, cex = 1, shape=21, col="black",
                      ggplot2::aes(x=X, y=Y, fill=grp), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill="Group")
```

@@

```{r}
#| label: clust-geo
#| echo: false

@@

clust_geo <- ClustGeo::hclustgeo(events_gower_mod,
                                 dist_geo, alpha = 0.1)
clust_geo_grp <- cutree(clust_geo, k=K)

>>>>

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
backup_options <- options()
options(scipen = 1, digits = 2)
do_eval <- FALSE
source(here::here("scripts", "R", "postprocess", "3_ordination.R"))
```

@@

```{r echo=FALSE}
#| label: load-data
anns_wide <- readRDS(here::here("data", "objects", "annotations_wide_treated.rds"))
anns_wide_noempty  <- readRDS(here::here("data", "objects", "annotations_wide_noempty.rds"))
dets <- readRDS(here::here("data", "objects", "detections.rds"))
exif <- readr::read_csv(here::here("data", "tabular", "exif_processed.csv"))
cams_sf_coords <- readRDS(here::here("data", "objects", "cams_sf_coords.rds"))
zones_filtered  <- readRDS(here::here("data", "objects", "zones_filtered.rds"))
zones_pols <- readRDS(here::here("data", "objects", "zones_pols.rds"))
base_map <- readRDS(here::here("data", "objects", "base_map.rds"))
camera_days <- readRDS(here::here("data", "objects" ,"camera_days.rds"))
camera_regions <- readRDS(here::here("data", "objects" ,"camera_regions.rds"))
extr_table_sub_tb <- readRDS(here::here("data", "objects" ,"extr_table_sub_tb.rds"))
anns_wide_animals_joined_mut <- readRDS(here::here("data", "objects", "annotations_animals.rds"))
anns_wide_animals_table_tb <- readRDS(here::here("data", "objects" ,"anns_wide_animals_table_tb.rds"))
events_tb_filt_with_col <- readRDS(here::here("data", "objects" ,"events_tb_filt_with_col.rds"))
events_cam_days <- readRDS(here::here("data", "objects" ,"events_cam_days.rds"))
events_tb_cam <- readRDS(here::here("data", "objects" ,"events_tb_cam.rds"))
sci_names_processed <- readRDS(here::here("data", "objects" ,"sci_names_processed.rds"))
table_cat <- readRDS(here::here("data", "objects" ,"table_cat.rds"))
```

@@

```{r echo=FALSE}
#| label: make-labels

regions_labels <- zones_filtered |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.5),
                              y = c(52, 52.6)))
```

```{r echo=FALSE}
#| label: fig-deployment
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."

p <- base_map +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords,
                      shape = 24, col = "black", fill = "red",
                      ggplot2::aes(x = X, y = Y), 
                      inherit.aes = FALSE)

ggplot2::ggsave(plot = p, filename = here::here("figures", "camera_array.png"), 
                width = 6, height = 8, dpi = 300)

# saveRDS(p, here::here("data", "objects", "camera_array.rds"))
# p <- readRDS(here::here("data", "objects", "camera_array.rds"))

p
```

@@

```{r echo=FALSE}
#| label: basic-stats-zero

@@

```{r echo=FALSE}
#| label: manual-annotations

@@

We summarized our existing species data to date for analysis in two main ways:  (1) species presence/absence data for each camera and, (2) relative species abundance, measured as the number of days, within each month that each species was detected at a camera site. The detection time for species is highly variable.

```{r echo=FALSE}
#| label: camera-plots
#| fig-cap: "Box plot of detection time for each species in either ecozone."
#| fig.width: 8
#| fig.height: 4.5

bp_data <-  anns_wide_animals_joined_mut |> 
  dplyr::left_join(camera_days, by = "camera_id") |>

@@

  dplyr::mutate(diff_time = difftime(first_day, camera_first_day, units = "days"),
                ECOZONE_NA = as.factor(ECOZONE_NA))

bp <- bp_data |> 
  ggplot2::ggplot(ggplot2::aes(x = diff_time, 
                               y = forcats::fct_reorder(species, total_events),
                               color = ECOZONE_NA)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_x_continuous() +
  ggplot2::labs(x = "Detection time (in days)",
                y = "Species",
                color = "Ecozone") + 
  ggplot2::theme_bw()

ggplot2::ggsave(plot = bp, filename = "figures/bp.png", width = 10, height = 15)
```

```{r echo=FALSE}
#| label: camera-plots-2
#| fig-cap: "Bubble plot of detection time for each species in either ecozone"
#| fig.width: 8
#| fig.height: 4.5

pts_data <- anns_wide_animals_joined_mut |>
  dplyr::left_join(camera_days, by = "camera_id") |>

@@

  dplyr::mutate(diff_time = difftime(first_day, camera_first_day, units = "days"),
                ECOZONE_NA = as.factor(ECOZONE_NA))

pts <- pts_data |> 
  ggplot2::ggplot(ggplot2::aes(x = diff_time, 
                               y = forcats::fct_reorder(species, total_events),
                               # size = camera_events, 
                               color = ECOZONE_NA)) +
  ggplot2::geom_point(ggplot2::aes(x = diff_time, 
                                   y = forcats::fct_reorder(species, total_events),
                                   size = camera_events, 
                                   color = ECOZONE_NA,
                                   group = ECOZONE_NA)) +
  ggplot2::scale_x_continuous() +
  ggplot2::labs(x = "Detection time (in days)",
                y = "Species",
                color = "Ecozone",
                size = "Day counts") + 
  ggplot2::theme_bw()

ggplot2::ggsave(plot = pts, filename = "figures/pts.png", width = 10, height = 15)

pts
```

@@

@@

```{r echo=FALSE}
#| label: tab-species-table

sp_tab <- dplyr::rename(.data = events_tb_cam |> colSums() |> 
                          sort(decreasing = TRUE) |> as.data.frame(), 
                        days_detected = `sort(colSums(events_tb_cam), decreasing = TRUE)`) |> 
  tibble::rownames_to_column("species")

sp_tab |>
  dplyr::left_join(sci_names_processed, by = "species") |> 
  dplyr::relocate(Scientific_name, .after = species) |> 
  dplyr::filter(days_detected >= 10) |>
  gt::gt() |>
  gt::tab_header(
    title = "Table 2: Relative abundances for species with more than 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) |>
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) |> 
  gt::cols_label(
    days_detected = "Days",
    species = "Species",
    Scientific_name = "Scientific name"
  ) |> 
  gt::tab_style(style = list(
      gt::cell_text(style = "italic")),
    locations = gt::cells_body(columns = Scientific_name)) |> 
  gt::opt_stylize(style = 6, color = "gray")

@@

```{r echo=FALSE}
#| label: fig-map-pies
#| fig-cap: "Pie chart map of relative species abundances across our study area Detections have grouped by plot for better visibility, with a plot containing up to 4 cameras. Only species detected more than 50 days in the season are shown for clarity."
#| fig.width: 8
#| fig.height: 6

cams_sf_with_props <- cams_sf_coords |> 
  sf::st_drop_geometry() |> 
  dplyr::select(plot_id,X,Y) |> 
  dplyr::group_by(plot_id) |> 
  dplyr::mutate(X = mean(X), Y = mean(Y)) |>
  dplyr::distinct() |> 
  dplyr::ungroup() |> 
  dplyr::left_join(events_tb_filt_with_col, by = "plot_id")

regions_labels_bis <- zones_filtered |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-85, -88),
                              y = c(52, 52.6)))

base_map +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x = X, y = Y, group = plot_id),
                              cols = names(cams_sf_with_props)[4:(4 + ncol(cams_sf_with_props) - 4)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x = "Lat", y = "Long") +
    ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels_bis)
```

@@

@@

```{r echo=FALSE}
#| label: rda-pa
#| echo: false

anns_wide_animals_table_tb_ch <- vegan::decostand(anns_wide_animals_table_tb, 
                                                  "hellinger")
pa_rda <- vegan::rda(anns_wide_animals_table_tb ~ .,
                     extr_table_sub_tb)
```

@@

```{r echo=FALSE}
#| label: fig-rda-pa
#| fig-cap: "Triplot of wildlife presence-absence data from Northern Ontario camera trapping images collected in March through September of 2022 reveals that different species associate with different land cover classes. For example, caribou cranes are more likely to be detected near open and treed wetlands (OBOG, OFEN, TrBOG, and TrFEN) while black bear and moose near forested areas (MixTRE, SpTRE, ConTRE)." 
custom_rda_plot(pa_rda)
```

@@

```{r echo=FALSE}
#| label: rda-events
#| fig-cap: "Triplot of RDA on relative abundance data"

@@

```{r echo=FALSE}
#| label: fig-rda-events
#| fig-cap: "Triplot of wildlife relative abundance data from Northern Ontario camera trapping images collected in March through September of 2022, showing similar patterns than the triplot using presence-absence data."

@@

# vegan::RsquareAdj(events_rda)
custom_rda_plot(events_rda, sp_scale = 1.5, scale_x_sp = 0.4, scale_y_sp = 0.3,
                thres_plot = 0.3)
```

@@

```{r echo=FALSE}
#| label: cluster-base
#| echo: false

@@

```{r echo=FALSE}
#| label: fig-rda-events-clust
#| fig-cap: "Triplot of wildlife relative abundance data from Northern Ontario camera trapping images collected in March through September of 2022, showing similar patterns than the triplot using presence-absence data. Two clusters determined using Ward clustering are shown as ellipses."

clust_grp <- cutree(clust, k = 4)
custom_rda_plot(events_rda, sp_scale = 1.5, scale_x_sp = 0.4, scale_y_sp = 0.3,
                thres_plot = 0.2716, clust = clust_grp)
```

@@

```{r echo=FALSE}
#| label: fig-map-clust
#| fig-cap: "Map of clustered camera sites. The two clusters were determined using the Ward clustering method."

@@

base_cut <- base_map +
  ggplot2::coord_map(xlim = c(-89, -84),
                     ylim = c(50, 54.5))

base_cut +
  ggplot2::geom_point(data = cams_sf_clust, cex = 1, shape = 21, col = "black",
                      ggplot2::aes(x = X, y = Y, fill = grp), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill = "Group")
```

@@

```{r echo=FALSE}
#| label: clust-geo

@@

clust_geo <- ClustGeo::hclustgeo(events_gower_mod,
                                 dist_geo, alpha = 0.1)
clust_geo_grp <- cutree(clust_geo, k = K)
